market-justice-preferences

### 2024-04-25 paper.tex
update references

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdfauthor={Equipo EDUMER},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}

\section{The socialization of meritocracy and market justice preferences
at
school}\label{the-socialization-of-meritocracy-and-market-justice-preferences-at-school}

\subsection{Introduction}\label{introduction}

Since its origins, educational institutions have been related to the
idea of social mobility and access to better opportunities. Therefore,
the consistent evidence of the high level of social reproduction at the
school level represents a threat to the promise of education and a
meritocratic system (\citeproc{ref-bourdieu_reproduction_1990}{Bourdieu
and Passeron 1990}). A large part of the research in this field at an
international level has addressed the extent to which the social origin
of students affects their academic results and their life opportunities
(\citeproc{ref-vonhippel_test_2019}{Von Hippel and Hamrock 2019}),
confirming that schools have severe difficulties in closing the gaps of
origin. Besides this socioeconomic perspective on school opportunities,
recent research has addressed to what extent inequalities in the school
context are also influencing students' perceptions, beliefs, and
attitudes: Are social inequalities perceived at the school context? Are
they rejected by the students, particularly those who are worst-off in
socioeconomic terms? Or, Is there evidence at the school level that
social inequalities are tolerated and even justified?
(\citeproc{ref-batruch_belief_2022}{Batruch et al. 2022};
\citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al. 2015}).~

In the present paper, we deal with the justification of social
inequalities by eighth-grade students in Chile, a country characterized
by a highly stratified educational system. In particular, we focus on
market justice preferences (Lane, 1986), which refer to the preferences
for distributing public goods (as health and education) based on
criteria such as competition and payment capacity. Although from a
rational point of view it could be expected an opposition to market
justice by the underprivileged majority,~ we argue that in a social
environment characterized by a promotion of meritocratic ideals - as the
school system - would lead to the opposite: a larger market justice
preferences.~

Given that the school environment has an important focus on performance,
achievement and acknowledgment, meritocracy has been one of the
principal concepts used for understanding and even for justifying
performance differences among students. Meritocracy is a distributive
system based on the belief that people should be rewarded and promoted
based on their abilities, knowledge, and achievements
(\citeproc{ref-young_rise_1958}{Young 1958}). It is often seen as a way
to create equal opportunities and fairness, as individuals can rise to
positions of power and influence based on their own merit rather than
their background or connections. However, some argue that meritocracy
can actually lead to tolerating or even justifying social inequalities,
as it can create a hierarchy where those who already have resources and
advantages are more likely to succeed. In this regard, a great deal of
academic research about meritocracy delves into the assessment of to
what extent rewards and privileges in society are related to merit,
emphasizing the so-called unfulfillable promise of meritocracy
(\citeproc{ref-mijs_stratified_2016}{J. Mijs 2016}). Complementing this
agenda, a second and emerging research area deals with subjective
aspects of meritocracy, such as perceptions and beliefs.~

@@

The perception of meritocracy refers to how individuals view and
understand the concept of meritocracy in their own society
(\citeproc{ref-duru-bellat_who_2012}{Duru-Bellat and Tenret 2012};
\citeproc{ref-castillo_meritocracia_2019}{Juan C. Castillo et al.
2019}). This perception can vary greatly depending on individual
experiences, social, economic, and cultural background. Some people may
see meritocracy as a fair and just system that allows anyone to succeed
based on their abilities and hard work. In contrast, others may view it
as a myth or a cover for existing power dynamics and inequality, serving
to maintain and even reinforce inequality
(\citeproc{ref-lampert_meritocratic_2013}{Lampert 2013};
\citeproc{ref-mijs_paradox_2021}{J. J. B. Mijs 2021}). Based in this
last perspective, we argue that individuals with a higher perception of
meritocracy will show a larger justification of social inequalities, as
individual achievement would be seen as rewarded and social policies as
less necessary (\citeproc{ref-batruch_belief_2022}{Batruch et al.
2022}).

Most of the research that has related meritocratic beliefs to inequality
justification so far has only considered adults, leaving aside the study
of how beliefs in this field develop at student age as well as the
impact of the school context and the family as the main socialization
agencies. Regarding schools, the way in which they deal with unequal
conditions of origin has been linked to the \emph{hidden curriculum}
(\citeproc{ref-chafel_schooling_1997}{Chafel 1997}), whereby students
learn about distributive norms in society and mechanisms of
justification of social differences. Based on recent studies that relate
school meritocracy to the justification of economic inequalities in the
adult population (\citeproc{ref-batruch_belief_2022}{Batruch et al.
2022}; \citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al. 2015}),
the central hypothesis guiding this research is that school-age students
with a higher perception of meritocracy - both at school and at the
societal level - will show a larger justification of social
inequalities, as individual achievement would be seen as appropriately
rewarded and social mechanisms for correcting inequalities as less
necessary (\citeproc{ref-batruch_belief_2022}{Batruch et al. 2022}).

@@

\subsection{Justification of
inequality}\label{justification-of-inequality}

Research on social stratification beliefs, which explore individual
perceptions of who deserves what and why
(\citeproc{ref-kluegel_beliefs_1987}{Kluegel and Smith 1987}),
highlights that people's explanations and justifications of social
inequality are closely tied to their judgments of deservingness. The
influence of ideologies (\citeproc{ref-wegener_dominant_1995}{Wegener
and Liebig 1995}) and cultural schemas
(\citeproc{ref-homan_being_2017}{Homan, Valentino, and Weed 2017}) is
pivotal in shaping these explanations by offering symbolic
representations that frame societal structures and expectations. While
significant attention has been paid to wage inequality, income
distribution, and payment differentials in the literature {[}Juan Carlos
Castillo (\citeproc{ref-castillo_legitimacy_2011}{2011}); {]} (Castillo,
2011; Evans et al., 2010; Jasso, 1999; Shariff et al., 2016),~ there has
been less examination of public beliefs about which life domains should
be governed by market relations and even less about children's
acceptance or rejection of these market principles. This oversight is
notable given the extensive encroachment of market logic into public
goods, welfare policy, and social services over the past five decades
(Centeno \& Cohen, 2012; Harvey, 2007), affecting areas such as
pensions, health services, and education. The justification of social
inequality based on marketype criteria has been conceptualized as the
individuals' adherence to one specific justice evaluation, the market
justice, that is, affording legitimacy to the allocation of goods and
services based on prices and individuals' ability to pay (Boltanski \&
Chiapello- 2007; Lane 1986; Streeck 2012).

Robert E. Lane proposed the underpinnings of market justice, which he
differentiated from political justice. For him, ``it is the genius of
the market to stimulate wants without at the same time stimulating a
sense of deserving more than one gets'' (1986, p.~384). Following the
theory of relative deprivation --a social phenomenon arising when
individuals cannot afford what most others in their environment can
(Merton 1950)-- Lane notes that, in market settings, social comparisons
are more likely to motivate increased effort rather than feelings of
acute injustice because individuals attribute outcomes to their actions.
Although empirical research has shown that, contrary to Lane's
observation, relative deprivation, even in market settings, produces
feelings of dissatisfaction, anger, and resentment that might motivate
forms of collective action such as protests and revolt (Greitemeyer and
Sagioglou 2016; Mishra and Carleton 2015; Smith et al.~2012; Power 2018;
Power, Madsen, and Morton 2020), the conceptualization of market justice
has been since then closely coupled to the merit principle for
allocating outcomes: the claim that the unequal levels of well-being
individuals enjoy ought to be, to some extent, a function of their
talents and efforts, regardless of their needs or membership, the two
latter being the realm of political justice and its closely coupled
principles of need --allocating outcomes to those who require them
most-- and equality --allocating the same outcome to everyone--, which
have been at the center of welfarism (see Wilson 2003). The underlying
notion of market justice also resonates in its application to welfare
regimes, providing a framework for understanding the varied global
approaches to managing social services.

The management of social services manifests in varying approaches across
nations, with substantial differences in funding and delivery methods
(Jensen, 2008; Stoy, 2014). Nordic countries, for example, predominantly
employ public agencies to produce and provide social services, funding
these through collective taxation and offering them in kind to the
majority of citizens. This system prioritizes political justice, placing
it above market mechanisms in accessing services. In contrast, other
countries rely more heavily on for-profit entities and private funding,
where service distribution depends mainly on individual financial
capacity of paying user fees, highlighting the influence of market
justice in service allocation. The trend toward marketization of welfare
services has been growing since the 1980s (Salamon, 1993), and this
shift is increasingly evident even in countries where market solutions
have traditionally had a minor role in social policy (Sivesind, 2017).

The question arises whether adults and children justify unequal access
to welfare services based on market justice principles. Influenced by
theories of policy feedback, which suggest that social welfare policies
can reinforce (positive feedback) or undermine (negative feedback)
previous policy trajectories
(\citeproc{ref-fernandez_positive_2013}{Fernandez and Jaime-Castillo
2013}) (Fernandez and Jaime-Castillo 2013; Pierson 2000; Weaver 2010),
citizens' beliefs about market justice are likely also shaped by the
institutional and social contexts they encounter. Indeed, the
justification of inequality in access to essential services like
education and health, based on one's ability to pay, shows significant
variation across countries, as demonstrated by international surveys,
although they do not usually include children in their samples. For
instance, the 2019 International Social Survey Program (ISSP) provides
insights into how adults perceive inequality in accessing welfare
services. Figure 1 illustrates the variation in agreement levels
regarding whether it is just for individuals with higher incomes to
purchase better education or healthcare. Notably, in 18 of the 29
countries analyzed, there is a greater justification for market
inequality in healthcare access than in education. Nevertheless, the
general sentiment typically ranges from `somewhat unjust' to 'neither
just nor unjust, with mixed feelings.

While contemporary policy developments have increasingly embraced the
marketization of social welfare in areas such as education, pensions,
and health services, questions about public acceptance of these changes
persist. Lindh (\citeproc{ref-lindh_public_2015}{2015}) analysis of ISSP
2009 data from 17 OECD countries reveals a general lack of support for
market-based distribution of social services, suggesting widespread
disapproval of market stratification of essential services. This finding
is corroborated by Soler-Martínez and colleagues' (2023) research from
Latinobarómetro 2020 across 18 Latin American countries, where concerns
about health and education access predominated over income inequality.
These results indicate that reforms toward welfare marketization are
typically driven by elite political decisions rather than grassroots
demand.~

Despite high-income inequality and limited social mobility in Latin
America, there is a prevalent belief that individuals are solely
responsible for their economic outcomes, a view that varies across the
region
(\citeproc{ref-buccaMeritBlameUnequal2016}{\textbf{buccaMeritBlameUnequal2016?}};
\citeproc{ref-chongMysteryDiscriminationLatin2008}{\textbf{chongMysteryDiscriminationLatin2008?}};
\citeproc{ref-torcheIntergenerationalMobilityInequality2014}{\textbf{torcheIntergenerationalMobilityInequality2014?}};
\citeproc{ref-salgadoInequalityStratificationLatin2023}{\textbf{salgadoInequalityStratificationLatin2023?}}).
The reliance on private welfare providers and widespread user fees
(Molyneux, 2008) adds complexity to this context. Yet, research on
children's justification of market-based inequalities in accessing
welfare services remains limited, especially in Latin America,
highlighting a significant gap in understanding how younger generations
view market-based access to welfare and whether these views are
associated with their meritocratic beliefs.~

\subsection{Meritocracy}\label{meritocracy}

The concept of meritocracy frequently appears nowadays when analyzing
cultural determinants of social inequalities. In general, it is
mentioned as a value associated with justice, as it would link efforts
and talents with rewards in an equitable manner. This normative sense is
quite far from its original formulation by Young
(\citeproc{ref-young_rise_1958}{1958}) in the satirical novel ``The Rise
of Meritocracy'', where it ironically represented a mechanism for
reproducing the inequalities of origin. The meritocratic ideal had
remained relatively unchallenged until a series of recent publications
turned into its potential consequences for maintaining social
inequality. Perhaps one of the most recent sources in this line is
Michael Sandel's ``The Tyranny of Merit'', where he strongly questions
the implications of carrying out the principle of merit in societies
that do not guarantee equal opportunities and that generate a feeling of
scarce recognition and appreciation of those who receive lesser rewards:
``In society's eyes, and perhaps also their own, their work no longer
signified a valued contribution to the common good.''
(\citeproc{ref-sandel_tyranny_2020}{Sandel 2020}, pp.).

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdfauthor={Equipo EDUMER},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

\hypertarget{the-socialization-of-meritocracy-and-market-justice-preferences-at-school}{%
\section{The socialization of meritocracy and market justice preferences
at
school}\label{the-socialization-of-meritocracy-and-market-justice-preferences-at-school}}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Since its origins, educational institutions have been related to the
idea of social mobility and access to better opportunities. Therefore,
the consistent evidence of the high level of social reproduction at the
school level represents a threat to the promise of education and a
meritocratic system
(\protect\hyperlink{ref-bourdieu_reproduction_1990}{Bourdieu and
Passeron 1990}). A large part of the research in this field at an
international level has addressed the extent to which the social origin
of students affects their academic results and their life opportunities
(\protect\hyperlink{ref-vonhippel_test_2019}{Von Hippel and Hamrock
2019}), confirming that schools have severe difficulties in closing the
gaps of origin. Besides this socioeconomic perspective on school
opportunities, recent research has addressed to what extent inequalities
in the school context are also influencing students' perceptions,
beliefs, and attitudes: Are social inequalities perceived at the school
context? Are they rejected by the students, particularly those who are
worst-off in socioeconomic terms? Or, Is there evidence at the school
level that social inequalities are tolerated and even justified?
(\protect\hyperlink{ref-batruch_belief_2022}{Batruch et al. 2022};
\protect\hyperlink{ref-wiederkehr_belief_2015}{Wiederkehr et al.
2015}).~

In the present paper, we deal with the justification of social
inequalities by eighth-grade students in Chile, a country characterized
by a highly stratified educational system. In particular, we focus on
market justice preferences
(\protect\hyperlink{ref-lane_market_1986}{Lane 1986}), which refer to
the preferences for distributing public goods (as health and education)
based on criteria such as competition and payment capacity. Although
from a rational point of view it could be expected an opposition to
market justice by the underprivileged majority,~ we argue that in a
social environment characterized by a promotion of meritocratic ideals -
as the school system - would lead to the opposite: a larger market
justice preferences.~

Given that the school environment has an important focus on performance,
achievement and acknowledgment, meritocracy has been one of the
principal concepts used for understanding and even for justifying
performance differences among students. Meritocracy is a distributive
system based on the belief that people should be rewarded and promoted
based on their abilities, knowledge, and achievements
(\protect\hyperlink{ref-young_rise_1958}{Young 1958}). It is often seen
as a way to create equal opportunities and fairness, as individuals can
rise to positions of power and influence based on their own merit rather
than their background or connections. However, some argue that
meritocracy can actually lead to tolerating or even justifying social
inequalities, as it can create a hierarchy where those who already have
resources and advantages are more likely to succeed. In this regard, a
great deal of academic research about meritocracy delves into the
assessment of to what extent rewards and privileges in society are
related to merit, emphasizing the so-called unfulfillable promise of
meritocracy (\protect\hyperlink{ref-mijs_stratified_2016}{Mijs 2016}).
Complementing this agenda, a second and emerging research area deals
with subjective aspects of meritocracy, such as perceptions and
beliefs.~

@@

The perception of meritocracy refers to how individuals view and
understand the concept of meritocracy in their own society
(\protect\hyperlink{ref-duru-bellat_who_2012}{Duru-Bellat and Tenret
2012}; \protect\hyperlink{ref-castillo_meritocracia_2019}{Castillo et
al. 2019}). This perception can vary greatly depending on individual
experiences, social, economic, and cultural background. Some people may
see meritocracy as a fair and just system that allows anyone to succeed
based on their abilities and hard work. In contrast, others may view it
as a myth or a cover for existing power dynamics and inequality, serving
to maintain and even reinforce inequality
(\protect\hyperlink{ref-lampert_meritocratic_2013}{Lampert 2013};
\protect\hyperlink{ref-mijs_paradox_2021}{Mijs 2021}). Based in this
last perspective, we argue that individuals with a higher perception of
meritocracy will show a larger justification of social inequalities, as
individual achievement would be seen as rewarded and social policies as
less necessary (\protect\hyperlink{ref-batruch_belief_2022}{Batruch et
al. 2022}).

Most of the research that has related meritocratic beliefs to inequality
justification so far has only considered adults, leaving aside the study
of how beliefs in this field develop at student age as well as the
impact of the school context and the family as the main socialization
agencies. Regarding schools, the way in which they deal with unequal
conditions of origin has been linked to the \emph{hidden curriculum}
(\protect\hyperlink{ref-chafel_schooling_1997}{Chafel 1997}), whereby
students learn about distributive norms in society and mechanisms of
justification of social differences. Based on recent studies that relate
school meritocracy to the justification of economic inequalities in the
adult population (\protect\hyperlink{ref-batruch_belief_2022}{Batruch et
al. 2022}; \protect\hyperlink{ref-wiederkehr_belief_2015}{Wiederkehr et
al. 2015}), the central hypothesis guiding this research is that
school-age students with a higher perception of meritocracy - both at
school and at the societal level - will show a larger justification of
social inequalities, as individual achievement would be seen as
appropriately rewarded and social mechanisms for correcting inequalities
as less necessary (\protect\hyperlink{ref-batruch_belief_2022}{Batruch
et al. 2022}).

@@

\hypertarget{justification-of-inequality}{%
\subsection{Justification of
inequality}\label{justification-of-inequality}}

Research on social stratification beliefs, which explore individual
perceptions of who deserves what and why
(\protect\hyperlink{ref-kluegel_beliefs_1987}{Kluegel and Smith 1987}),
highlights that people's explanations and justifications of social
inequality are closely tied to their judgments of deservingness. The
influence of ideologies
(\protect\hyperlink{ref-wegener_dominant_1995}{Wegener and Liebig 1995})
and cultural schemas (\protect\hyperlink{ref-homan_being_2017}{Homan,
Valentino, and Weed 2017}) is pivotal in shaping these explanations by
offering symbolic representations that frame societal structures and
expectations. While significant attention has been paid to wage
inequality, income distribution, and payment differentials in the
literature {[}Castillo
(\protect\hyperlink{ref-castillo_legitimacy_2011}{2011}); Evans et al.,
2010; Jasso (\protect\hyperlink{ref-jasso_how_1999}{1999}) ; Shariff,
Wiwad, and Aknin
(\protect\hyperlink{ref-shariff_income_2016}{2016}){]},~ there has been
less examination of public beliefs about which life domains should be
governed by market relations and even less about children's acceptance
or rejection of these market principles. This oversight is notable given
the extensive encroachment of market logic into public goods, welfare
policy, and social services over the past five decades
(\protect\hyperlink{ref-centeno_arc_2012}{Centeno and Cohen 2012};
\protect\hyperlink{ref-harvey_breve_2015}{Harvey 2015}), affecting areas
such as pensions, health services, and education. The justification of
social inequality based on marketype criteria has been conceptualized as
the individuals' adherence to one specific justice evaluation, the
market justice, that is, affording legitimacy to the allocation of goods
and services based on prices and individuals' ability to pay
(\protect\hyperlink{ref-boltanski_new_2005}{Boltanski and Chiapello
2005}; \protect\hyperlink{ref-lane_market_1986}{Lane 1986};
\protect\hyperlink{ref-streeck_citizens_2012}{Streeck 2012}).

Robert E. Lane proposed the underpinnings of market justice, which he
differentiated from political justice. For him, ``it is the genius of
the market to stimulate wants without at the same time stimulating a
sense of deserving more than one gets'' {[}-Lane
(\protect\hyperlink{ref-lane_market_1986}{1986}); p.~384{]}. Following
the theory of relative deprivation --a social phenomenon arising when
individuals cannot afford what most others in their environment can
(Merton 1950)-- Lane notes that, in market settings, social comparisons
are more likely to motivate increased effort rather than feelings of
acute injustice because individuals attribute outcomes to their actions.
Although empirical research has shown that, contrary to Lane's
observation, relative deprivation, even in market settings, produces
feelings of dissatisfaction, anger, and resentment that might motivate
forms of collective action such as protests and revolt
(\protect\hyperlink{ref-greitemeyer_subjective_2016}{Greitemeyer and
Sagioglou 2016}; \protect\hyperlink{ref-mishra_subjective_2015}{Mishra
and Carleton 2015}; \protect\hyperlink{ref-smith_relative_2012}{Smith et
al. 2012}; \protect\hyperlink{ref-power_deprivationprotest_2018}{Séamus
A. Power 2018}; \protect\hyperlink{ref-power_relative_2020}{Séamus A.
Power, Madsen, and Morton 2020}), the conceptualization of market
justice has been since then closely coupled to the merit principle for
allocating outcomes: the claim that the unequal levels of well-being
individuals enjoy ought to be, to some extent, a function of their
talents and efforts, regardless of their needs or membership, the two
latter being the realm of political justice and its closely coupled
principles of need --allocating outcomes to those who require them
most-- and equality --allocating the same outcome to everyone--, which
have been at the center of welfarism (see Wilson 2003). The underlying
notion of market justice also resonates in its application to welfare
regimes, providing a framework for understanding the varied global
approaches to managing social services.

The management of social services manifests in varying approaches across
nations, with substantial differences in funding and delivery methods
(\protect\hyperlink{ref-jensen_worlds_2008}{Jensen 2008};
\protect\hyperlink{ref-stoy_worlds_2014}{Stoy 2014}). Nordic countries,
for example, predominantly employ public agencies to produce and provide
social services, funding these through collective taxation and offering
them in kind to the majority of citizens. This system prioritizes
political justice, placing it above market mechanisms in accessing
services. In contrast, other countries rely more heavily on for-profit
entities and private funding, where service distribution depends mainly
on individual financial capacity of paying user fees, highlighting the
influence of market justice in service allocation. The trend toward
marketization of welfare services has been growing since the 1980s
(\protect\hyperlink{ref-salamon_marketization_1993}{Salamon 1993}), and
this shift is increasingly evident even in countries where market
solutions have traditionally had a minor role in social policy
(\protect\hyperlink{ref-sivesind_changing_2017}{Sivesind 2017}).

The question arises whether adults and children justify unequal access
to welfare services based on market justice principles. Influenced by
theories of policy feedback, which suggest that social welfare policies
can reinforce (positive feedback) or undermine (negative feedback)
previous policy trajectories
(\protect\hyperlink{ref-fernandez_positive_2013}{Fernandez and
Jaime-Castillo 2013};
\protect\hyperlink{ref-pierson_increasing_2000}{Pierson 2000};
\protect\hyperlink{ref-weaver_paths_2010}{Weaver 2010}), citizens'
beliefs about market justice are likely also shaped by the institutional
and social contexts they encounter. Indeed, the justification of
inequality in access to essential services like education and health,
based on one's ability to pay, shows significant variation across
countries, as demonstrated by international surveys, although they do
not usually include children in their samples. For instance, the 2019
International Social Survey Program (ISSP) provides insights into how
adults perceive inequality in accessing welfare services. Figure 1
illustrates the variation in agreement levels regarding whether it is
just for individuals with higher incomes to purchase better education or
healthcare. Notably, in 18 of the 29 countries analyzed, there is a
greater justification for market inequality in healthcare access than in
education. Nevertheless, the general sentiment typically ranges from
`somewhat unjust' to 'neither just nor unjust, with mixed feelings.

While contemporary policy developments have increasingly embraced the
marketization of social welfare in areas such as education, pensions,
and health services, questions about public acceptance of these changes
persist. Lindh (\protect\hyperlink{ref-lindh_public_2015}{2015})
analysis of ISSP 2009 data from 17 OECD countries reveals a general lack
of support for market-based distribution of social services, suggesting
widespread disapproval of market stratification of essential services.
This finding is corroborated by Soler-Martínez, García-Sánchez, and
Willis (\protect\hyperlink{ref-soler-martinez_concerns_2023}{2023})
research from Latinobarómetro 2020 across 18 Latin American countries,
where concerns about health and education access predominated over
income inequality. These results indicate that reforms toward welfare
marketization are typically driven by elite political decisions rather
than grassroots demand.~

Despite high-income inequality and limited social mobility in Latin
America, there is a prevalent belief that individuals are solely
responsible for their economic outcomes, a view that varies across the
region (\protect\hyperlink{ref-bucca_merit_2016}{Bucca 2016};
\protect\hyperlink{ref-chong_mystery_2008}{Chong and Ñopo 2008};
\protect\hyperlink{ref-torche_intergenerational_2014}{Torche 2014};
\protect\hyperlink{ref-salgado_inequality_2023}{Salgado and Castillo
2023}). The reliance on private welfare providers and widespread user
fees (\protect\hyperlink{ref-molyneux_neoliberal_2008}{Molyneux 2008})
adds complexity to this context. Yet, research on children's
justification of market-based inequalities in accessing welfare services
remains limited, especially in Latin America, highlighting a significant
gap in understanding how younger generations view market-based access to
welfare and whether these views are associated with their meritocratic
beliefs.~

\hypertarget{meritocracy}{%
\subsection{Meritocracy}\label{meritocracy}}

The concept of meritocracy frequently appears nowadays when analyzing
cultural determinants of social inequalities. In general, it is
mentioned as a value associated with justice, as it would link efforts
and talents with rewards in an equitable manner. This normative sense is
quite far from its original formulation by Young
(\protect\hyperlink{ref-young_rise_1958}{1958}) in the satirical novel
``The Rise of Meritocracy'', where it ironically represented a mechanism
for reproducing the inequalities of origin. The meritocratic ideal had
remained relatively unchallenged until a series of recent publications
turned into its potential consequences for maintaining social
inequality. Perhaps one of the most recent sources in this line is
Michael Sandel's ``The Tyranny of Merit'', where he strongly questions
the implications of carrying out the principle of merit in societies
that do not guarantee equal opportunities and that generate a feeling of
scarce recognition and appreciation of those who receive lesser rewards:
``In society's eyes, and perhaps also their own, their work no longer
signified a valued contribution to the common good.''
(\protect\hyperlink{ref-sandel_tyranny_2020}{Sandel 2020}, pp.).
